---
title: "Case_study_on_Divvy_Bikes"
author: "Satyendra"
date: '2022-03-28'
output: html_document
---
##Business Task-

Understand how casual riders and annual members of Cyclistic - Bike Share use Cyclistic bikes differently and come up with efficient recommendations to convert casual riders into annual members

##Stakeholders-

  1-Lily Moreno - Director of Marketing
  2-Cyclistic marketing analytics team
  3-Cyclistic executive team

##Ask Phase of Data Analysis

Notes:As it is clear that we want to convert casual members into annual                     members, we can come up with these SMART questions -

    *How do annual members and casual riders use Cyclistic bikes differently?
    *Why would casual riders buy Cyclistic annual memberships?
    *How can Cyclistic use digital media to influence casual riders to become       members?

#key Question
“In what ways do members and casual riders use Divvy bikes                differently?”

## Setting up the environment
Notes:Importing required libraries
```{r }
library(tidyverse)
```
```{r}
library(ggplot2)
```
```{r}
library(lubridate) # for time and date 
```

## Importing the dataset of Divvy bikes for year 2021
Notes: Data of every month

```{r }
mon_jan<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202101-divvy-tripdata.csv")
```
```{r}
mon_feb<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202102-divvy-tripdata.csv")
```
```{r}
mon_march<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202103-divvy-tripdata.csv")
```
```{r}
mon_april<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202104-divvy-tripdata.csv")
```
```{r}
mon_may<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202105-divvy-tripdata.csv")
```
```{r}
mon_june<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202106-divvy-tripdata.csv")
```
```{r}
mon_july<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202107-divvy-tripdata.csv")
```
```{r}
mon_aug<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202108-divvy-tripdata.csv")
```
```{r}
mon_sept<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202109-divvy-tripdata.csv")
```
```{r}
mon_oct<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202110-divvy-tripdata.csv")
```
```{r}
mon_nov<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202111-divvy-tripdata.csv")
```
```{r}
mon_dec<-read.csv("D:\\google cert\\case\\zip files\\unzip files\\202112-divvy-tripdata.csv")
```

## Getting glimpse of column names of data

```{r}
colnames(mon_jan)
```
Notes:Every month has same no of columns
```{r}
colnames(mon_july)
```
```{r}
colnames(mon_dec)
```

##Understanding the structure of data -for January month(similar to each month)

```{r}
str(mon_jan)
```
###feb montth

```{r}
str(mon_feb)
```
```{r}
str(mon_july)
```
```{r}
str(mon_dec)
```

Notes: From structure it is clear that the data type of start_station_id and end_station_id does not have consistent type

## Making changes in data types of specific columns of each month

```{r}
mon_jan<-  mutate(mon_jan, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id)) 
```
```{r}
mon_feb<-  mutate(mon_feb, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```
```{r}
mon_march<-mutate(mon_march, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```
```{r}
mon_april<-mutate(mon_april, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```
```{r}
mon_may<-mutate(mon_may, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```
```{r}
mon_june<-  mutate(mon_june, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```
```{r}
mon_july<-  mutate(mon_july, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```
```{r}
mon_aug<-  mutate(mon_aug, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```
```{r}
mon_sept<-  mutate(mon_sept, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```
```{r}
mon_oct<-  mutate(mon_oct, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```
```{r}
mon_nov<-  mutate(mon_nov, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```
```{r}
mon_dec<-  mutate(mon_dec, start_station_id = as.character(start_station_id)
                      ,end_station_id = as.character(end_station_id))
```


#Combining all month data into a single dataset 

```{r}
trips<-bind_rows(mon_jan,mon_feb,mon_march,mon_april,mon_may,mon_june,mon_july,mon_aug,mon_sept,mon_oct,mon_nov,mon_dec) 
```
```{r}
str(trips)
```
# DATA CLEANING
Notes :dropping column of no use
```{r}
trips<-trips %>% 
  select(-c(start_lat,start_lng,end_lat,end_lng))
str(trips)
```
Notes:Details of Members of Divvy Bikes
```{r}
table(trips$member_casual)
```
Notes: Changing names from casual to "customer" and memeber to "Subscriber"
```{r}
trips<-trips %>% 
  mutate(member_casual=recode(member_casual,
                              "member"="Subscriber","casual"="Customer"))
str(trips)
```
```{r}
table(trips$member_casual) # total no of customers of particular type
```
##Getting Glimpse of modified data 
```{r}
colnames(trips)
nrow(trips) # no of rows in complete data set
```
```{r}
dim(trips) #dimension of combined data set
```
```{r}
head(trips) # head rows of trips dataset 
```
```{r}
summary(trips) #generating summary 
```

##Adding new columns
```{r}
trips$date<-as.Date(trips$started_at) #extracting date from started date column format yyyy-mm-dd
```
```{r}
trips$month<-format(as.Date(trips$date),"%m") #extracting month from date column
```
```{r}
trips$day <- format(as.Date(trips$date), "%d")#extracting day from date
```
```{r}
trips$year <- format(as.Date(trips$date), "%Y")#extracting year from date
```
```{r}
trips$day_of_week <-format(as.Date(trips$date),"%A")#extracting day of week from date
```

##Calculating ride length of each customer for our analysis
```{r}
trips$ride_length<-difftime(trips$ended_at,trips$started_at) #ended_at and started_at columns difference give ride length of the rider in seconds
```
```{r}
head(trips,5)
```


Notes:Changing type of ride_length column
```{r}
is.factor(trips$ride_length)
```

```{r}
trips$ride_length <- as.numeric(as.character(trips$ride_length))
```

#Analytical phase 
##Generating summary with respect to ride_length column
```{r}
summary(trips$ride_length)
```

Notes:since negative values are present so we remove them
```{r}
trips_v2 <- trips[!(trips$start_station_name == "HQ QR" | trips$ride_length<0),]
```

##Generating summary again
```{r}
summary(trips$ride_length)
```

Notes:Mean of ride length with respect to membership status 
```{r}
aggregate(trips_v2$ride_length ~ trips_v2$member_casual, FUN = mean)
```
Notes:Median of ride length with respect to membership status
```{r}
aggregate(trips_v2$ride_length ~ trips_v2$member_casual, FUN = median)
```

Notes:Max of ride length with respect to membership status
```{r}
aggregate(trips_v2$ride_length ~ trips_v2$member_casual, FUN = max)
```

Notes:Min of ride length with respect to membership status
```{r}
aggregate(trips_v2$ride_length ~ trips_v2$member_casual, FUN = min)
```
```{r}
#mean of data with respect to week days and membership status 
aggregate(trips_v2$ride_length ~ trips_v2$member_casual + trips_v2$day_of_week,
          FUN = mean)
```
```{r}
# analyze ridership data by type and weekday
trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual,weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```

```{r}
# calculates the average duration by month
trips_v2 %>%
  group_by(month, member_casual) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(month, member_casual)
```

##DATA VISUALIZATION
```{r}
#plotting data --memmbership v/s day wise
trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(weekday,member_casual) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge", width = 0.8) +
 theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Average duration of rides in particular week days")
  
```


```{r}
#plotting monthly wise--month v/s membership
trips_v2 %>%
  group_by(month, member_casual) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(month, member_casual) %>%
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge", width = 0.5) +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Average duration of rides from Jan 2021 to Dec 2021")
```

```{r}
#plotting with respect to ridertype
trips_v2 %>%
  group_by(rideable_type, member_casual) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(rideable_type, member_casual) %>%
  ggplot(aes(x = rideable_type, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge", width = 0.5) +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Average duration of rides from Jan 2021 to Dec 2021")
```


```{r}
counts <- aggregate(trips_v2$ride_length ~ trips_v2$member_casual +
                      trips_v2$day_of_week, FUN = mean)
```
```{r}
counts
```
                     
                     